Tool-driven Revolutions in Archaeological Science

### 2019-08-12 analysis/paper/jcaa-paper.Rmd
ideas-driven and tools-driven to idea-driven and tool-driven

In this paper we investigate recent developments which are facilitating computational archaeology's return to fundamental principles of the scientific method. We ask whether this might be the beginning of a broader change in the field. Revolutions or paradigm-shifts in archaeology have been proposed to occur following major theoretical statements [@Clark_1993; @Härke_2002], but there is an argument among philosophers of science whether revolutions in science are ideas-driven or tools-driven [@Galison_1997; @Dyson_2000]. This argument motivated us to explore how the disciplines of ecology and archaeology have started to embrace tools that improve the reproducibility of research. We focus on one set of tools for this change, based on writing data analysis scripts in free and open source programming languages, exemplified by R, and the practice of sharing of these scripts. We focus on R because our prior observations indicate it is by far the most common scripting language used by ecologists and archaeologists. Bibliometric evidence shows a strong increase in the use of R among ecologists and the beginning of a similar development in archaeology. We evaluate how approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and propose these might be a tool-driven change in archaeology. Acknowledging that this is not a simple process, we offer an R-based tool to ease the task of creating a compendium which enables other researchers to reproduce the published results.

# Revolutions in Science: Ideas-driven or Tools-driven?

Philosophers of science disagree whether revolutions in science are ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of 'revolution'. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another. In this conversion or speciation process the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, as it describes a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of evolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm, but a conceptual vacuum [@wilkins1996are].

>>>>

In this paper we investigate recent developments which are facilitating computational archaeology's return to fundamental principles of the scientific method. We ask whether this might be the beginning of a broader change in the field. Revolutions or paradigm-shifts in archaeology have been proposed to occur following major theoretical statements [@Clark_1993; @Härke_2002], but there is an argument among philosophers of science whether revolutions in science are idea-driven or tool-driven [@Galison_1997; @Dyson_2000]. This argument motivated us to explore how the disciplines of ecology and archaeology have started to embrace tools that improve the reproducibility of research. We focus on one set of tools for this change, based on writing data analysis scripts in free and open source programming languages, exemplified by R, and the practice of sharing of these scripts. We focus on R because our prior observations indicate it is by far the most common scripting language used by ecologists and archaeologists. Bibliometric evidence shows a strong increase in the use of R among ecologists and the beginning of a similar development in archaeology. We evaluate how approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and propose these might be a tool-driven change in archaeology. Acknowledging that this is not a simple process, we offer an R-based tool to ease the task of creating a compendium which enables other researchers to reproduce the published results.

# Revolutions in Science: idea-driven or tool-driven?

Philosophers of science disagree whether revolutions in science are idea-driven or tool-driven. One of the most widely-known models of idea-driven change in science is Kuhn's [-@Kuhn_1962] effort to describe the history of science by proposing paradigm shifts resulting from revolutionary change in communities of researchers. In brief, this model proposes that the history of science consists of long periods of tradition-bound 'normal science' punctuated by short episodes of 'revolution'. Normal science was described by Kuhn as stagnant, routine day-to-day research focused on 'puzzle solving'. What counts as suitable puzzles, and acceptable solutions to these puzzles, is governed by the norms and procedures of the prevailing paradigm. Revolutions in science occur as punctuations in the equilibria of normal science, when anomalous observations culminate in changes replacing one paradigm with another. In this conversion or speciation process the community adopts a new way of doing science. There has been much debate about how to draw a distinction between so-called normal and revolutionary science. [@watkins1970against; @toulmin1970does; @Casadevalle00158_16]. A notable criticism is that Kuhn's normal science is unscientific, as it describes a situation where critical science had contracted into defensive metaphysics resulting from the domination of a ruling dogma [@popper1970]. This is at odds with a view of science as a continuous state of evolution, with researchers simultaneously employing many styles of thinking and doing science, and continuously confronting theory with evidence and modifying their ideas based on the outcome. In this view, based on the post-World War II era of science, radical discontinuities are rare, and major developments instead emerge from the division and recombination of already mature fields [@andersen2013second]. Given the rarity of radical discontinuities, some biologists have argued that change in biology is better characterised by new ideas (e.g. Mendelian heredity, Darwinian evolution, and molecular genetics) replacing not a former paradigm, but a conceptual vacuum [@wilkins1996are].
